An inside-outside algorithms for mutant RNAs with applications to error detections in structured RNAs

### 2012-10-05 Recomb/introduction_RECOMB.tex
RNApyro

In recent years, studies as the \emph{Human Microbiome Project}~\cite{Turnbaugh2007}, 
leveraging the NGS techniques to sequence as many new organisms 
as possible, are producing a wealth of new information. Although
those techniques have a huge throughput, they yield a sequencing error rate of around
$4\%$~\cite{Huse2007}. This error can be highly reduced in genome projects when highly 
redundant consensus 
assemblies are available, but in studies of new or not well known organisms, there is not
 enough  similarity to differentiate between the sequencing errors and the natural 
 polymorphisms that we want to observe, often inflating the diversity estimates~\cite{Kunin2010}.
  In rRNAs, we have as additional 
 information the conserved secondary structure, and we want to use it to identify
highly probable sequencing errors.

The first challenge is to efficiently explore the mutant space, which grows exponentially. 
Leveraging the techniques  in \texttt{RNAmutants}~\cite{Waldispuhl2008}, we develop 
here

>>>>

In recent years, studies as the \emph{Human Microbiome Project}~\cite{Turnbaugh2007}, 
leveraging the NGS techniques to sequence as many new organisms 
as possible, are producing a wealth of new information. Although
those techniques have a huge throughput, they yield a sequencing error rate of around
$4\%$~\cite{Huse2007}. This error can be highly reduced  when highly 
redundant multiple sequence allignments 
 are available, but in studies of new or not well known organisms, there is not
 enough  similarity to differentiate between the sequencing errors and the natural 
 polymorphisms that we want to observe, often inflating the diversity estimates~\cite{Kunin2010}.
  In rRNAs, on top of multiple sequence alignments, we have as additional 
 information the conserved secondary structure, and we want to use it to identify
highly probable sequencing errors. 
Given a new rRNA sequence, a first challenge to find the sequencing errors
  is to  efficiently explore the mutant space, 
which grows exponentially, guiding ourselves with the alignment and the consensus structure.

Leveraging the techniques  in \texttt{RNAmutants}~\cite{Waldispuhl2008}, and building on top 
of the \emph{Inside-Outside algorithm}, we define here a new method called \texttt{RNApyro}
efficiently computing for large RNAs those probabilities.
 Classical techniques define a probabilistic model using a Boltzmann distribution 
whose weights are based on the free energy of the structure, using as energy parameter
the values of Turner found in the NNDB~\cite{Turner2010}  for stacked, 
canonical and wobble, base pairs. As shown by Leontis and Westhof~\cite{Leontis2001},
this  does not encapsulated the large diversity of base pairs that any nucleotide
can form with any other, although with an energy too small to be yet determined
by experimental techniques. To quantify geometrical differences, they
 define an isostericity distance, increasing as two base pairs differ 
 more from one another in space. We incorporate this second measure in the Boltzmann weights.

To benchmark

### 2012-10-05 Recomb/methods_RECOMB.tex
Quelques typos et suppression des fichiers temporaires
fixed typos
RNApyro

\subsubsection{Energy contribution}
The free-energy contribution in our pseudo-energy model corresponds to an additive stacking-pairs model, taking values from the Turner \todo[fancyline]{which version of the Turner model?}XXXXX model retrieved from the NNDB~\cite{Turner2010}. Given a candidate sequence $s$ for a secondary structure $S$, the free-energy of $S$ on $s$ is given by
\begin{align*}
  \ES(s,S) = \sum_{\substack{(i,j)\to (i',j')\in S\\ \text{stacking pairs}}}\ES^{\beta}_{s_is_j\to s_{i'}s_{j'}} 
\end{align*}
where $\ES^{\beta}_{ab\to a'b'}$ is the tabulated free-energy of a stacking pair in the Turner model if available, $0$ if $ab=\varnothing$ (no base-pair to stack onto), and using default value $\beta\in[0,\infty]$ for non-Watson-Crick/Wobble entries (i.e. not in $\{\text{GU},\text{UG},\text{CG},\text{GC}, \text{AU or UA}\}$). This latter parameter allows to choose whether to completely forbid a sequence inducing invalid base pairs ($\beta = \infty$), or to simply penalize it.
The imprecision introduced by this simplification of the Turner model remains reasonable, since the targeted secondary structure is fixed 
(e.g. multiloops do not account for base-specific contributions). Furthermore, it greatly eases the design and implementation of dynamic-programming equations. 
\subsection{isostericity contribution}
The concept of isostericity captures the geometric superimposability of two base-pairs, using individual scores given by Stombaugh~\emph{et al}~\cite{Stombaugh2009} based on a probabilistic model. Let $s$ be a candidate sequence for a secondary structure $S$, given in the context of a gapless RNA alignment $\Omega$,  the contribution of the isostericity to the pseudo-energy is given by
\begin{align*}
  \ES(s,S,\Omega) &= \sum_{\substack{(i,j)\in S\\ \text{pairs}}}\EI^{\Omega}_{(i,j),s_i s_j}, & \text{where}&& 	\EI^{\Omega}_{(i,j),ab}:=
	\frac{
		\sum_{s'\in\Omega}
			\text{\ISO}((s_i',s_j'),(a,b))}
%-		\left(			\text{\ISO}((s_i',s_j'),(s_i,s_j))				\right)	
{		
		|\Omega|
	}
\end{align*}
is the average isostericity of a base-pair in the candidate sequence in comparison with the reference alignment.
The $\ISO$ function uses the \todo[fancyline, inline]{\relsize{-1}Is this true? If so, we might want to try taking the most likely edge/orientation combination for each couple instead.}{Watson-Crick/Watson-Crick} cis isostericity matrix computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Isostericity scores range between $0$ and $9.7$, with $0$ being assigned to a perfect isostericity, and a penalty of $10$ is used for missing entries.
The isostericity contribution will favor exponentially sequences that are likely to adopt a similar local conformation as the alignment sequences.

@@

\subsubsection{Outside}	
The \emph{Outside} function, $\mathcal Y$, is the partition function considering only the 
energy in subsequences $[0,i]\cup[j,n-1]$ over the mutants of $s$ having exactly $m$ mutations between $[0,i]\cup[j,n-1]$ and whose nucleotide at position $i+1$ is $a$ 
(resp. in position $j-1$ it is $b$).
We define function $\Y{i,j}{m}{a,b}$ as a recurrence, and will use as initial conditions:
\begin{equation}
	\Y{-1,j}{m}{X,X}:=
		\displaystyle
	  \Z{j,n-1}{m}{X,X}
\label{eq:Y_in}
\end{equation}
The recurrence, as shown below, will increase the interval $[i,j]$ by decreasing $i$ when
it is not base paired. If it is with a position $k>j$, we increase $j$ to include it.
 Thus, when we need
to evaluate an interval as $(-1,j)$, all stems between $(0,j)$ are taken into account and the
structure between $(j,n-1)$ must be a set of independent stems. Therefore,
 all the outside energy between $[j,n-1]$ is
equal to $\Z{j,n-1}{m}{X,X}$, for any $X\in B$. The recursion itself is as follows.
\begin{equation}
	\Y{i,j}{m}{a,b} = \left\{
  \begin{array}{ll}
		\displaystyle
    \sum_{\substack{a'\in \B,\\ \Kron_{a',s_i}\le m}}
    \Y{i-1,j}{m- \Kron_{a',s_i}}{a',b} &
    \text{Elif }S_i=-1 \\
    \displaystyle
    \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_is_j}\le m}}
		 e^{\frac{-E_{(i,j),ab \to a'b'}^{\Omega,\beta}}{RT}}\cdot
    \Y{i-1,j+1}{m- \Kron_{a'b',s_is_j}}{a',b'} &
   	 \text{Elif }S_{i}=j \land S_{i+1}=j-1\\
		 \displaystyle
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_is_k}\le m}}
		 \sum_{m'=0}^{m-\Kron_{a'b',s_is_k}}
  		 e^{\frac{E_{(i,k),\varnothing\to a'b'}^{\Omega,\beta}}{RT}}
		 \cdot\Y{i-1,k+1}{m- \Kron_{a'b',s_is_k} - m'}{a',b'}
     \cdot\Z{j,k-1}{m'}{b,b'} &
		 \text{Elif }S_{i}=k \geq j\\
		 \displaystyle
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_ks_i}\le m}}
		 \sum_{m'=0}^{m-\Kron_{a'b',s_ks_i}}
   	 e^{\frac{E_{(k,i),\varnothing\to a'b'}^{\Omega,\beta}}{RT}}
		 \cdot\Y{k-1,j}{m- \Kron_{a'b',s_ks_i} - m'}{a',b}
     \cdot\Z{k+1,i-1}{m'}{a',b'} &
		 \text{Elif }-1 < S_{i}=k < i\\
		 0 & \text{Else}
  \end{array}\right.
\label{eq:Y_rec}
\end{equation}
The five cases can be broked down as follows.
\begin{description}
\item[$S_i=-1$:] If the nucleotide at position $i$ is not paired, then the value is the same
as if we decrease the lower interval bound by $1$ (i.e. $i-1$), and consider all possible
nucleotides $a'$ at position $i$, correcting the number of mutants
in function of $\Kron_{a',s_i}$.
\item[$S_{i}=j$ and $S_{i+1}=j-1$:] If nucleotide $i$ is paired with $j$ and nucleotide $i+1$ is
paired with $j-11$, we are in the only case were stacked base pairs can occur. We thus add
the energy of the stacking and of the isostericity of the base pair $(i,j)$. What is left
to compute is the \emph{outside} value for the interval $[i-1,j+1]$ over all possible nucleotides 
$a',b'\in B^2$ at positions $i$ and $j$ respectively.
\item[$S_{i}=k \geq j$:]If nucleotide $i$ is paired with position $k\geq j$, 
and is not stacked inside, the 
only term contributing directly to the energy is the isostericity of the base pair $(i,k)$. 
We can then consider the outside interval $[i-1,k+1]$ by multiplying it by the the \emph{forward}
value of the newly included interval (i.e. $[j,k-1]$), for 
all possible values $a',b'\in B^2$ for nucleotides at positions $i$ and $k$ respectively.
\item[$-1<S_{i}<i$:]As above but if position $i$ is to pairing with a lower value.
\item[Else:] In all other cases, we are in a derivation of the SCFG that does not correspond to the 
secondary structure $S$, and we return $0$.

@@

We are now left to compute the probability that a \todo[fancyline]{Maybe worth doing it for base-pairs? maybe not\ldots}{given position is a given nucleotide}.
We leverage the \emph{Inside-Outside} construction to immediately obtain the following $3$ cases.
Given $i\in[0,n-1],x\in B$, and $M\geq 0$ a bound on the number of allowed mutations. 
\begin{align*}
	\mathbb{P}(s_i = x\mid s,\Omega, S,M) &:= \frac{\mathcal{B}(i,x,s,\Omega,S,M)}{\sum_{m=0}^{M}\Z{0,n-1}{m}{X,X}}\\ 
\mathcal{B}(i,x,s,\Omega,S,M)&=
 \left\{
	\begin{array}{ll}
			\sum_{m=0}^{M}
			\Y{i-1,i+1}{m-\Kron_{x,s_i}}{x,x}
		&\text{If }S_i = -1\\
			\sum_{m=0}^{M}
			\sum_{\substack{b\in B\\\Kron_{xb,s_is_k\leq m}}}
			\sum_{m'=0}^{m-\Kron_{xb,s_is_k}}
     	 e^{\frac{E_{(i,k),\varnothing\to xb}^{\Omega,\beta} }{RT}}
			\Y{i-1,k+1}{m-\Kron_{xb,s_is_k-m'}}{x,b}
			\Z{i+1,k-1}{m'}{x,b}
		&\text{If }S_i=k>i\\
			\sum_{m=0}^{M}
			\sum_{\substack{b\in B\\\Kron_{bx,s_ks_i\leq m}}}
			\sum_{m'=0}^{m-\Kron_{bx,s_ks_i}}
     	 e^{E_{(k,i),\varnothing\to bx}{RT}}
			\Y{k-1,i+1}{m-\Kron_{bx,s_ks_i-m'}}{b,x}
			\Z{k+1,i-1}{m'}{b,x}
		&\text{If }S_i=k<i
	\end{array}\right.
\end{align*}

>>>>

\subsubsection{Energy contribution}
The free-energy contribution in our pseudo-energy model corresponds to an additive stacking-pairs model, taking values from the Turner 2004 model retrieved from the NNDB~\cite{Turner2010}. Given a candidate sequence $s$ for a secondary structure $S$, the free-energy of $S$ on $s$ is given by
\begin{align*}
  \ES(s,S) = \sum_{\substack{(i,j)\to (i',j')\in S\\ \text{stacking pairs}}}\ES^{\beta}_{s_is_j\to s_{i'}s_{j'}} 
\end{align*}
where $\ES^{\beta}_{ab\to a'b'}$ is the tabulated free-energy of a stacking pair in the Turner model if available, $0$ if $ab=\varnothing$ (no base-pair to stack onto), and using default value $\beta\in[0,\infty]$ for non-Watson-Crick/Wobble entries (i.e. not in $\{\text{GU},\text{UG},\text{CG},\text{GC}, \text{AU or UA}\}$). This latter parameter allows to choose whether to completely forbid a sequence inducing invalid base pairs ($\beta = \infty$), or to simply penalize it.
The imprecision introduced by this simplification of the Turner model remains reasonable, since the targeted secondary structure is fixed 
(e.g. multiloops do not account for base-specific contributions). Furthermore, it greatly eases the design and implementation of dynamic-programming equations. 
\subsection{isostericity contribution}
The concept of isostericity captures the geometric superimposability of two base-pairs, using individual scores given by Stombaugh~\emph{et al}~\cite{Stombaugh2009} based on a probabilistic model. Let $s$ be a candidate sequence for a secondary structure $S$, given in the context of a gapless RNA alignment $\Omega$,  the contribution of the isostericity to the pseudo-energy is given by
\begin{align*}
  \ES(s,S,\Omega) &= \sum_{\substack{(i,j)\in S\\ \text{pairs}}}\EI^{\Omega}_{(i,j),s_i s_j}, & \text{where}&& 	\EI^{\Omega}_{(i,j),ab}:=
	\frac{
		\sum_{s'\in\Omega}
			\text{\ISO}((s_i',s_j'),(a,b))}
%-		\left(			\text{\ISO}((s_i',s_j'),(s_i,s_j))				\right)	
{		
		|\Omega|
	}
\end{align*}
is the average isostericity of a base-pair in the candidate sequence in comparison with the reference alignment.
The $\ISO$ function uses the \todo[fancyline, inline]{\relsize{-1}Is this true? If so, we might want to try taking the most likely edge/orientation combination for each couple instead.}{Watson-Crick/Watson-Crick} cis isostericity matrix computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Isostericity scores range between $0$ and $9.7$, with $0$ being assigned to a perfect isostericity, and a penalty of $10$ is used for missing entries.
The isostericity contribution will favor exponentially sequences that are likely to adopt a similar local conformation as the alignment sequences.

@@

\subsubsection{Outside}	
The \emph{Outside} function, $\mathcal Y$, is the partition function considering only the 
energy in subsequences $[0,i]\cup[j,n-1]$ over the mutants of $s$ having exactly $m$ mutations between $[0,i]\cup[j,n-1]$ and whose nucleotide at position $i+1$ is $a$ 
(resp. in position $j-1$ it is $b$).
We define function $\Y{i,j}{m}{a,b}$ as a recurrence, and will use as initial conditions:
\begin{equation}
	\Y{-1,j}{m}{X,X}:=
		\displaystyle
	  \Z{j,n-1}{m}{X,X}
\label{eq:Y_in}
\end{equation}
The recurrence, as shown below, will increase the interval $[i,j]$ by decreasing $i$ when
it is not base paired. If it is with a position $k>j$, we increase $j$ to include it.
 Thus, when we need
to evaluate an interval as $(-1,j)$, all stems between $(0,j)$ are taken into account and the
structure between $(j,n-1)$ must be a set of independent stems. Therefore,
 all the outside energy between $[j,n-1]$ is
equal to $\Z{j,n-1}{m}{X,X}$, for any $X\in B$. The recursion itself is as follows.
\begin{equation}
	\Y{i,j}{m}{a,b} = \left\{
  \begin{array}{ll}
		\displaystyle
    \sum_{\substack{a'\in \B,\\ \Kron_{a',s_i}\le m}}
    \Y{i-1,j}{m- \Kron_{a',s_i}}{a',b} &
    \text{Elif }S_i=-1 \\
    \displaystyle
    \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_is_j}\le m}}
		 e^{\frac{-E_{(i,j),ab \to a'b'}^{\Omega,\beta}}{RT}}\cdot
    \Y{i-1,j+1}{m- \Kron_{a'b',s_is_j}}{a',b'} &
   	 \text{Elif }S_{i}=j \land S_{i+1}=j-1\\
		 \displaystyle
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_is_k}\le m}}
		 \sum_{m'=0}^{m-\Kron_{a'b',s_is_k}}
  		 e^{\frac{-E_{(i,k),\varnothing\to a'b'}^{\Omega,\beta}}{RT}}
		 \cdot\Y{i-1,k+1}{m- \Kron_{a'b',s_is_k} - m'}{a',b'}
     \cdot\Z{j,k-1}{m'}{b,b'} &
		 \text{Elif }S_{i}=k \geq j\\
		 \displaystyle
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_ks_i}\le m}}
		 \sum_{m'=0}^{m-\Kron_{a'b',s_ks_i}}
   	 e^{\frac{-E_{(k,i),\varnothing\to a'b'}^{\Omega,\beta}}{RT}}
		 \cdot\Y{k-1,j}{m- \Kron_{a'b',s_ks_i} - m'}{a',b}
     \cdot\Z{k+1,i-1}{m'}{a',b'} &
		 \text{Elif }-1 < S_{i}=k < i\\
		 0 & \text{Else}
  \end{array}\right.
\label{eq:Y_rec}
\end{equation}
The five cases can be broked down as follows.
\begin{description}
\item[$S_i=-1$:] If the nucleotide at position $i$ is not paired, then the value is the same
as if we decrease the lower interval bound by $1$ (i.e. $i-1$), and consider all possible
nucleotides $a'$ at position $i$, correcting the number of mutants
in function of $\Kron_{a',s_i}$.
\item[$S_{i}=j$ and $S_{i+1}=j-1$:] If nucleotide $i$ is paired with $j$ and nucleotide $i+1$ is
paired with $j-11$, we are in the only case were stacked base pairs can occur. We thus add
the energy of the stacking and of the isostericity of the base pair $(i,j)$. What is left
to compute is the \emph{outside} value for the interval $[i-1,j+1]$ over all possible nucleotides 
$a',b'\in B^2$ at positions $i$ and $j$ respectively.
\item[$S_{i}=k \geq j$:]If nucleotide $i$ is paired with position $k\geq j$, 
and is not stacked inside, the 
only term contributing directly to the energy is the isostericity of the base pair $(i,k)$. 
We can then consider the outside interval $[i-1,k+1]$ by multiplying it by the the \emph{forward}
value of the newly included interval (i.e. $[j,k-1]$), for 
all possible values $a',b'\in B^2$ for nucleotides at positions $i$ and $k$ respectively.
\item[$-1<S_{i}<i$:]As above but if position $i$ is to pairing with a lower value.
\item[Else:] In all other cases, we are in a derivation of the SCFG that does not correspond to the 
secondary structure $S$, and we return $0$.

@@

We are now left to compute the probability that a \todo[fancyline]{Maybe worth doing it for base-pairs? maybe not\ldots}{given position is a given nucleotide}.
We leverage the \emph{Inside-Outside} construction to immediately obtain the following $3$ cases.
Given $i\in[0,n-1],x\in B$, and $M\geq 0$ a bound on the number of allowed mutations. 
\begin{align*}
	\mathbb{P}(s_i = x\mid s,\Omega, S,M) &:= \frac{\mathcal{B}(i,x,s,\Omega,S,M)}{\sum_{m=0}^{M}\Z{0,n-1}{m}{X,X}}\\ 
\mathcal{B}(i,x,s,\Omega,S,M)&=
 \left\{
	\begin{array}{ll}
			\sum_{m=0}^{M}
			\Y{i-1,i+1}{m-\Kron_{x,s_i}}{x,x}
		&\text{If }S_i = -1\\
			\sum_{m=0}^{M}
			\sum_{\substack{b\in B\\\Kron_{xb,s_is_k\leq m}}}
			\sum_{m'=0}^{m-\Kron_{xb,s_is_k}}
     	 e^{\frac{-E_{(i,k),\varnothing\to xb}^{\Omega,\beta} }{RT}}
			\Y{i-1,k+1}{m-\Kron_{xb,s_is_k-m'}}{x,b}
			\Z{i+1,k-1}{m'}{x,b}
		&\text{If }S_i=k>i\\
			\sum_{m=0}^{M}
			\sum_{\substack{b\in B\\\Kron_{bx,s_ks_i\leq m}}}
			\sum_{m'=0}^{m-\Kron_{bx,s_ks_i}}
     	 e^{\frac{-E^{\Omega, \beta}_{(k,i),\varnothing\to bx}}{RT}}
			\Y{k-1,i+1}{m-\Kron_{bx,s_ks_i-m'}}{b,x}
			\Z{k+1,i-1}{m'}{b,x}
		&\text{If }S_i=k<i
	\end{array}\right.
\end{align*}
